Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-08-03 submission/manuscript.Rmd
update abstract with Ada suggestions

Marc A Sze${^1}$, Nicholas A Lesniak${^1}$, Mack T Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

@@

**Background.** Colorectal cancer (CRC) is a continuing health concern with the majority of the risk for developing disease being due to environmental factors. The microbiota is one of these environmental factors with certain bacterial community members being associated with CRC and other taxa being associated to colons without tumors. Some of the bacterial species in taxa associated to colons without tumors can use fiber to produce short-chain fatty acids (SCFAs) that can inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in human CRC is less certain. Here, we test the hypothesis that SCFA concentrations are different in individuals with colorectal tumors.

**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their concentrations of acetate, butyrate, and propionate. Analysis also included imputed gene relative abundance with PICRUSt, metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group, and tumor classification and SCFA prediction models using Random Forest.

**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15). Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70). Finally, there was no difference between classification models with or without SCFAs in their ability to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).

**Conclusions.** Although our data does not support the hypothesis that SCFAs are different in individuals that have colorectal tumors, there may be context specific scenarios where SCFAs may still be beneficial for treatment of CRC. Alternatively, our observations also support the hypothesis that there may be other mechanisms that have not been thoroughly investigated that are more involved with the development of human CRC.

@@

**Total significant positive correlations between OTU relative abundance and SCFA concentration were similar for individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. Using Spearmans rho we found that the majority of significant OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5]. A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6]. There was a noticeablely higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas  [Figure 3]. Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

@@

The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors. Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2]. Although there were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have tumors, had an adenoma, or had a carcinoma, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1]. Instead, SCFA concentrations provided similar information to what specific OTUs were already providing to the tumor classification models [Figure 4]. Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to  these taxa in the tumor models [Figure 4-5]. Collectively, our observations suggest that resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae*, that are different between individuals without tumors and patients with adenomas or carinomas, are not that same as those involved with SCFA production.

Although SCFAs have been shown to be anti-tumorigenic, most of these studies have been performed in model systems [@test_OKeefe2016; @Encarnao2018]. Additionally, many of the *in vivo* studies investigate proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017]. Although it is well known that breakdown products from gut bacteria results in SCFA production, fiber effects on tumorigenesis may be through other mechanisms in these *in vivo* model systems. In addition to using fiber supplementation as a proxy for SCFA treatment, observations on the benefit of SCFAs in preventing tumorigenesis has been mixed. In previous case/control studies lower SCFA concentrations between patients with carcinomas and those without carcinomas was observed [@Ohigashi2013]. Yet, this is in contrast to multiple randomized-controlled trials that have found no difference between patients who do and do not get fiber supplementation to try and prevent tumor recurrence [@Schatzkin2000; @Yao2017]. This could be because fiber has other actions in humans besides providing an energy source for bacteria to breakdown to make SCFAs. It could also be due to the fact that SCFA concentrations and responses to fiber can vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016]. This information taken together with our observations would suggest that either indviduals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.

@@

One possible technical limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon. However, this is unlikely to be a major confounder. First, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018]. Second, previous studies that measure SCFA changes after fiber supplementation use fecal material to track these responses with a great deal of success [@scfa_measures_venkataraman2016]. Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be cause of this. Additionally, as mentioned earlier, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer. Yet, taxa that are associated with SCFA production are consistently higher in indivdiuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Indeed our data would support the contention that the taxa are similar to those associated with SCFA production but that these specific microbes or OTUs themselves are not associated with SCFAs. In particular, our results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models supports this hypothesis. This leads to the possibilty that protection may be through two other routes. First, there could be a different pathway or other less extensively studied metabolites that provides the necessary protection against tumorigenesis. Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016; @Kostic2011]). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics. Even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.

@@

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identifed and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].

@@

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of important mediators of SCFA synthesis was tested. Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify high or low SCFA concentration based on the median of each SCFA or the actual concentration using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in accuracy.

>>>>

Marc A. Sze${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

@@

**Background.** Colorectal cancer (CRC) is increasing in prevalence in individuals under 50 and because of this will be a continuing health concern for the foreseeable future. The majority of the risk for developing CRC is attributable to environmental factors. One of these environmental factors is the microbiota, with certain bacterial community members being associated with CRC and other taxa being associated to individuals without tumors. Some of the bacterial species in taxa associated to individuals without tumors can use fiber to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in human CRC is less certain. Here, we test the hypothesis that SCFA concentrations and taxa associated with their production are different in individuals with colorectal tumors.

**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their fecal concentrations of acetate, butyrate, and propionate. Analysis also included imputed gene relative abundance with PICRUSt, metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group, and tumor classification and SCFA prediction models using Random Forest.

**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15). Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70). Finally, there was no difference between the ability of Random Forest models to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).

**Conclusions.** Although our data does not support the hypothesis that fecal SCFA concentrations in patients in the general CRC population are different, there still may be specific types of colorectal tumors where SCFAs may be beneficial for treatment of CRC. Alternatively, our observations also support the hypothesis that there may be other metabolites or mechanisms (e.g. bacterial niche exclusion) that may be more protective against tumorigenesis and have not been thoroughly investigated in the context of human CRC.

@@

**Total significant positive correlations between OTU relative abundance and SCFA concentration were similar for individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. Using Spearmans rho we found that the majority of significant OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5]. A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6]. There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas  [Figure 3]. Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

@@

The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors. Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2]. Although there were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have tumors, had an adenoma, or had a carcinoma, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1]. Instead, SCFA concentrations provided similar information to what specific OTUs were already providing to the tumor classification models [Figure 4]. Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to  these taxa in the tumor models [Figure 4-5]. Collectively, our observations suggest that resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae*, that are different between individuals without tumors and patients with adenomas or carcinomas, are not that same as those involved with SCFA production.

Although SCFAs have been shown to be anti-tumorigenic, most of these studies have been performed in model systems [@test_OKeefe2016; @Encarnao2018]. Additionally, many of the *in vivo* studies investigate proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017]. Although it is well known that breakdown products from gut bacteria results in SCFA production, fiber effects on tumorigenesis may be through other mechanisms in these *in vivo* model systems. In addition to using fiber supplementation as a proxy for SCFA treatment, observations on the benefit of SCFAs in preventing tumorigenesis has been mixed. In previous case/control studies lower SCFA concentrations between patients with carcinomas and those without carcinomas was observed [@Ohigashi2013]. Yet, this is in contrast to multiple randomized-controlled trials that have found no difference between patients who do and do not get fiber supplementation to try and prevent tumor recurrence [@Schatzkin2000; @Yao2017]. This could be because fiber has other actions in humans besides providing an energy source for bacteria to breakdown to make SCFAs. It could also be due to the fact that SCFA concentrations and responses to fiber can vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016]. This information taken together with our observations would suggest that either individuals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.

@@

One possible technical limitation is that a fecal sample may not be an ideal type of bio-specimen and that the effect SCFAs have on tumorigenesis is only detected in the colon. However, this is unlikely to be a major confounder. First, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018]. Second, previous studies that measure SCFA changes after fiber supplementation use fecal material to track these responses with a great deal of success [@scfa_measures_venkataraman2016]. Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be cause of this. Additionally, as mentioned earlier, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer. Yet, taxa that are associated with SCFA production are consistently higher in individuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Indeed our data would support the contention that the taxa are similar to those associated with SCFA production but that these specific microbes or OTUs themselves are not associated with SCFAs. In particular, our results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models supports this hypothesis. This leads to the possibility that protection may be through two other routes. First, there could be a different pathway or other less extensively studied metabolites that provides the necessary protection against tumorigenesis. Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016; @Kostic2011]). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics. Even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.

@@

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].

@@

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of important mediators of SCFA synthesis was tested. Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify high or low SCFA concentration based on the median of each SCFA or the actual concentration using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.
